Estimating the number of clusters using Cross Validation

### 2015-06-23 cvclust.tex
minor formatting edits
cut out old well-separated cluster proof; simplify simulation results table

\documentclass[11pt]{article}
\usepackage{epsf}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{footnote}
\usepackage[bottom]{footmisc}
\usepackage[singlelinecheck=false]{caption}
\usepackage{caption}
\usepackage{amsthm}
\usepackage{enumitem}
\usepackage{graphics}
\usepackage{float}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{lscape}
\usepackage{array}
\textheight=8.7in
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{multirow}

@@

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}

@@

\section{Asymptotic Analysis of $k$-Means}

@@

@@

@@

\subsection{Well-separated clusters with noise}

Same with previous, the data set $\mathbf{X} \in \R^{n \times p}$ been
partitioned as $\mathbf{X}^L \in \R^{n \times L}$ and $\mathbf{X}^R
\in \R^{n \times R}$, with following set of assumptions:

\begin{enumerate}[label={Assumption},leftmargin=*]
	\item $1.$ There are $K$ true centers $(\mu_1,...,\mu_K)$ where $\mathbf{X}$ generated from
	\item $2.$ The $K$ partitioned centers $(\mu^L_1,...,\mu^L_K)$ are all distinct in $L$ dimensions: $||\mu^L_i-\mu^L_j|| \neq 0$ for any $i \neq j$; $(\mu^R_1,...,\mu^R_K)$ are all distinct in $R$ dimensions
	\item $3.$ There is noise, i.e. $$x_i = \mu \left(c_i\right) + \varepsilon_i, \hspace{0.2 in} \mu \left(c_i\right) \in \{\mu_1,...,\mu_K\}$$ where vector $\varepsilon_i$ is i.i.d with zero mean and covariance matrix $\Sigma$, which is independent from any $\mu_j \in (\mu_1,...,\mu_K)$, $c_i$ is the true cluster for observation $i$, which implies $$x^L_i = \mu^L \left(c_i\right) + \varepsilon^L_i,\hspace{0.2in} x^R_i = \mu^R \left(c_i\right) + \varepsilon^R_i$$ 
with $$\mu^L\left(c_i\right) \in \{\mu^L_1,...,\mu^L_K\},\hspace{0.2in} \mu^R\left(c_i\right) \in \{\mu^R_1,...,\mu^R_K\} $$\
where $\varepsilon^L_i$ and $\varepsilon^R_i$ are independent and both have zero mean and covariance matrix $\Sigma^L$ and $\Sigma^R$ respectively, so $\Sigma = 
\begin{pmatrix}
  \Sigma^L &0\\
  0  &\Sigma^R\\
  \end{pmatrix}$
. Also $\varepsilon^R_i$ is bounded $||\varepsilon^R_i|| \leq 1$.
	\item $4.$ Clusters are well separated, i.e. \[ ||x_i-\mu \left(c_i\right)|| < ||x_i-\mu_{j}|| \hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i \] and \[ ||x^L_i-\mu^L \left(c_i\right)|| < ||x^L_i-\mu^L_{j}||-\delta,  \hspace{0.2 in}||x^R_i-\mu^R \left(c_i\right)|| < ||x^R_i-\mu^R_{j}||-\delta \hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\] where $\delta > 0$ is a small constant 
	\item $5.$ The training set contains $n^{train}_j>0$ observations from each cluster $j$ and the test set contains $n^{test}_j > 0$ observations from each cluster $j$, $j=1,2...,K$
	\item $6.$ The prior for $x_i$ come from cluster $\mu_j$ is $\pi_j > 0$. So when the total number of observation in training set $m$ goes infinity, $\lim_{m\to \infty} n^{train}_j = m \cdot \pi_j $. Also $m=n\cdot r$, $n-m = n\cdot (1-r)$, where $0<r<1$ is a positive constant.
\end{enumerate}

\subsubsection{Consistent assumption for K-means}
\cite{pollard1981strong} found conditions that ensure the almost sure convergence of the set of cluster centers of K-means as the sample size increase. This section mainly introduce the theorem from \cite{pollard1981strong} and show the assumptions needed for our own theorem.\\\\  
Applying K-means clustering on i.i.d observations $x_1,x_2,...x_m$ from population $P$ with parameter $k$ is to choose cluster centers $a_1,a_2,...,a_k$ to minimize
\[	W_m = \frac{1}{m}\sum^m_{l=1}min_{1\leq j \leq k}||x_l-a_j||^2	\]
It's convenient to treat $W_m$ as a function of the set of cluster centers $A$ and of the  empirical distribution $P_m$ obtained from the sample by placing mass $m^{-1}$ at each of $x_1,x_2,...x_m$. Then the problem is to minimize 
\[ W(A,P_m):=\int min_{a\in A}||x-a||^2P_m(dx)\]
Over all possible choices of the set $A$ containing $k$ points. \\\\
For each probability measure $Q$ and each set $A$, defined
\[	\Phi(A,Q):= \int min_{a\in A}||x-a||^2 Q(dx)	\] and \[ \mathfrak{m}_k(Q):=inf\{\Phi(A,Q):A \hspace{0.05in} \text{contains} \hspace{0.05in} k  \hspace{0.05in} \text{or fewer points}\} \]
The theorem proved by \cite{pollard1981strong} says that
\begin{theorem}
	Suppose that $\int ||x||^2P(dx) < \infty$ and that for each $j=1,2,...,k$ there is a unique set $\bar{A}(j)$ for which $\Phi(\bar{A}(j),P)=\mathfrak{m}_j(P)$. Then $A_m \to \bar{A}(k)$ a.s and $\Phi(A_m,P_m) \to \mathfrak{m}_k(P)$ a.s 
\end{theorem}
\noindent
Where $A_m = A_m(k)$ is the set of optimal sample cluster centers satisfy $\Phi(A_m,P_m) = \mathfrak{m}_k(P_m)$ and $\bar{A}=\bar{A}(k)$ is the set of optimal population cluster center satisfy  $\Phi(\bar{A},P) = \mathfrak{m}_k(P)$. Such theorem gives the condition for the sample estimator of cluster centers be consistent estimator of optimal cluster centers.\\

In our setting, observations $x^R_1, x^R_2,...,x^R_m$ can be viewed as i.i.d observations from same distribution $\mathcal{P}$, which is a mixed distribution of $K$ distinct distributions with prior $\pi_1,\pi_2,...,\pi_K$.
\begin{theorem}
Suppose assumption $3$ holds and for each $j=1,2,...,k_{max}$ there is a unique set $\bar{A}(j)$ for which $\Phi(\bar{A}(j),\mathcal{P})=\mathfrak{m}_j(\mathcal{P})$, also $ \bar{A}(K) = \{\mu^R_1,\mu^R_2,...,\mu^R_K\}$. Then for any $k \leq k_{max}$, $A_m(k) \to \bar{A}(k)$ a.s.
\end{theorem}
\begin{proof}
By assumption $3$, $||\varepsilon^R_i||\leq 1$, which implies the first condition in theorem $2$ holds, i.e.
\[\int ||x||^2P(dx) < \infty\] 
Since the second condition of theorem $2$ is given as assumption in theorem $3$, we have almost sure convergence of cluster centers $A_m(k)$ for any $k \leq k_{max}$ by theorem $2$. 
\end{proof}
\noindent
In the rest of this section, the assumptions (conditions) given in theorem $3$ is called consistent assumptions. 
\subsubsection{Asymptotic property of Gabriel method}
\begin{definition}
Define event\\ $\mathcal{D} = \{\forall x_i \in \mathbf{X}, ||x^L_i-\hat{\mu}^L \left(c_i\right)|| < ||x^L_i-\hat{\mu}^L_{j}||, ||x^R_i-\hat{\mu}^R \left(c_i\right)|| < ||x^R_i-\hat{\mu}^R_{j}|| \hspace{0.2 in} \forall \hspace{0.01 in}j\neq c_i\}$, where $\hat{\mu}^R_{j}$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$and $\hat{\mu}^L_{j}$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^L_{train}$
\end{definition}

\begin{proposition}
Under assumptions $1-6$, $P(\mathcal{D}) \to 1$ with $m \to \infty$.
\end{proposition}
\begin{proof}
	By assumption $4$, \[  ||x^L_i-\mu^L \left(c_i\right)|| < ||x^L_i-\mu^L_{j}||-\delta \hspace{0.2in} \forall \hspace{0.01 in}j\neq c_i \] 
	Hence, 
	\begin{align} \nonumber
		||x^L_i-\hat{\mu}^L \left(c_i\right)|| & \leq \nonumber ||x^L_i-\mu^L \left(c_i\right)||+ ||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) ||\\ \nonumber
		&< ||x^L_i-\mu^L_{j}||-\delta + ||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) ||  \\ \nonumber
		& \leq ||x^L_i-\hat{\mu}^L_{j}||+||\mu^L_{j}-\hat{\mu}^L_{j}||-\delta + ||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) ||  \\ \nonumber
		& =  ||x^L_i-\hat{\mu}^L_{j}||-\left\lbrace \delta-||\mu^L_{j}-\hat{\mu}^L_{j}|| -||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) || \right\rbrace \\ 
m \to \infty		&<||x^L_i-\hat{\mu}^L_{j}||
	\end{align} 
The step $(5)$ is because for any given $\varepsilon > 0$, by Markov inequality
	\begin{align} \nonumber
	  P\left(||\mu^L_{j}-\hat{\mu}^L_{j}||^2 \geq \varepsilon \right) &\leq \frac{E||\mu^L_{j}-\hat{\mu}^L_{j}||^2}{\varepsilon}  \nonumber \\ \nonumber
	    &=  \frac{E\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)}{\varepsilon}  \\ \nonumber
	    &=  \frac{E[ tr\left\lbrace\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)\right\rbrace]}{\varepsilon}  \\ \nonumber
	   &=  \frac{E[ tr\left\lbrace\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\right\rbrace]}{\varepsilon}  \\ \nonumber
	  &=  \frac{ tr\left\lbrace E\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)\left(\mu^L_{j}-\hat{\mu}^L_{j}\right)^T\right\rbrace}{\varepsilon}  \\ \nonumber
	  &= \frac{ tr\{\Sigma^L \}}{n^{train}_j \cdot \varepsilon}   \\ \nonumber
	% \lim_{m\to \infty} n^{train}_j = m \cdot \pi_j \to \infty \hspace{1in} &\to 0  \\ \nonumber
	\end{align} 
So
\begin{align} \nonumber
  max_{j \in \{1,2,...,K\}} \hspace{0.05in} P\left(||\mu^L_{j}-\hat{\mu}^L_{j}||^2 \geq \varepsilon \right) &\leq max_{j \in \{1,2,...,K\}} \hspace{0.05in}\frac{E||\mu^L_{j}-\hat{\mu}^L_{j}||^2}{\varepsilon}  \nonumber \\ \nonumber
  &=max_{j \in \{1,2,...,K\}} \hspace{0.05in}  \frac{ tr\{\Sigma^L \}}{n^{train}_j \cdot \varepsilon}   \\ \nonumber
m \to \infty \hspace{0.3in}  &= \frac{ tr\{\Sigma^L \}}{m\hspace{0.02in}\pi^*\hspace{0.02in} \varepsilon}  \hspace{0.5in} \text{where}\hspace{0.1in}\pi^* = min_{j \in \{1,2,...,K\}}\pi_j\\ \nonumber
&\to 0
\end{align}

Therefore $\forall j, \hspace{0.05in} ||\mu^L_{j}-\hat{\mu}^L_{j}|| \to 0$ uniformly as $m \to \infty$. Hence, 
\begin{equation}
\lim_{m\to \infty} \left\lbrace \delta-||\mu^L_{j}-\hat{\mu}^L_{j}|| -||\mu^L \left(c_i\right)-\hat{\mu}^L \left(c_i\right) || \right\rbrace >0
\end{equation}

Same argument yields 
\[||x^R_i-\hat{\mu}^R \left(c_i\right)|| < ||x^R_i-\hat{\mu}^R_{j}||\]
Because $(6)$ holds uniformly $\forall j\in\{1,2,...,K\}$, we have 
\[ \lim_{m \to \infty}P(\mathcal{D})  =1\]
	\end{proof}
\begin{proposition}
Given assumption $1-6$ and the consistent assumptions of Theorem $3$, when $m \to \infty$, $P\left[A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}\right]=1$, where $\hat{\mu}^R_j$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$
\end{proposition}
\begin{proof}
By theorem $3$, we know $A_m(K) \to \bar{A}(K)$ a.s. and $\bar{A}(K) = \{\mu^R_1,\mu^R_2,...,\mu^R_K\}$. Denote $A_m(K) = \{a_1, a_2,..., a_K\}$, by almost sure convergence of $A_m(K)$ to $\bar{A}(K)$, we have\\\\
For all $\varepsilon > 0$, there exists $N_{\varepsilon}$ such that if $m > N_{\varepsilon}$, then
\[ P\left(||a_j-\mu^R_j||< \varepsilon\hspace{0.1in}\forall j\right)=1\]
 Now, take $\varepsilon = \frac{\delta}{2}$, then exists $N_{\delta}$ such that for all $m > N_{\delta}$
\[P\left(||a_j-\mu^R_j||<\frac{\delta}{2} \hspace{0.1in}\forall j\right)=1\]
By assumption $4$, 
\[||x^R_i-\mu^R \left(c_i\right)|| < ||x^R_i-\mu^R_{j}||-\delta \hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\]
Let $m > N_{\delta}$, for all $x^R_i \in \mathbf{X}^R_{train}$, with probability $1$

\begin{align}
  ||x^R_i - a(c_i)||&\leq ||x^R_i-\mu^R(c_i)|| + ||\mu^R(c_i)-a(c_i)||\nonumber \\ \nonumber
   &< ||x^R_i-\mu^R_{j}||-\delta + ||\mu^R(c_i)-a(c_i)|| \hspace{0.2 in} \forall \hspace{0.01 in}j\neq c_i  \\ \nonumber
    &\leq ||x^R_i-a_{j}|| + ||a_j - \mu^R_{j}||-\delta + ||\mu^R(c_i)-a(c_i)||  \\ \nonumber
 ||\mu^R(c_i)-a(c_i)||< \frac{\delta}{2} \hspace{0.05in}\& \hspace{0.05in}||a_j - \mu^R_{j}||< \frac{\delta}{2} \hspace{0.2in} &<||x^R_i-a_{j}||\hspace{0.2 in}  \forall \hspace{0.01 in}j\neq c_i 
\end{align}
Hence, $x^R_i$ is assigned to cluster center $a(c_i)$ by K-means. Note also that observations from same cluster are assigned to the same center. Since $a_j$ is the sample mean of all the observations assigned to it, we have
\[a_j = \hat{\mu}^R_j \hspace{0.1in}\forall j\]
 where $\hat{\mu}^R_j$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$. The proof is complete.
\end{proof}
\begin{proposition}
Conditioning on $\mathcal{D}$ and $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$, there exists a classification method works perfectly in case of $k = K$: the prediction error is $0$. That is, $\forall x_i \in \mathbf{X}_{test}$ prediction on $x^L_i$ always correctly predict cluster where $x^R_i$ is in. 
\end{proposition}

\begin{proof}
Given $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$,  we know from the proof of proposition $3$ that when applying K-means on $\mathbf{X}^R_{train}$ with parameter $K$, the solution of
\[ \underset{\left(\hat{\mu}^R\left(g_1\right),...,\hat{\mu}^R\left(g_K\right)\right)}{\operatorname{argmin}} \sum^K_{j=1} \sum_{g_j = \hat{g} \left(x^R_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 \]
is 
\begin{equation}
\{\hat{\mu}^R\left(g_j\right) = \hat{\mu}^R_{j}, j=1,...,K\}
\end{equation}
with 
\begin{equation}
\hat{g} \left(x^R_i\right) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
\end{equation}
where $\hat{g} \left(x^R_i\right)$ is the assigned cluster by K-means for observation $x^R_i \in \mathbf{X}^R_{train}$  , $\hat{\mu}^R\left(g_j\right)$ is the sample mean (centroid) of K-means returned cluster $g_j$, $\hat{\mu}^R_{j}$ is the sample mean of $j^{th}$ cluster in $\mathbf{X}^R_{train}$, $c_i$ is the true cluster for observation $i$.\\\\
The training set of $\mathcal{C}$ consists of data in the form of 
\[
\left(\mathbf{X}^L_{train},\mathcal{G} \right) = \left\lbrace \left(x^L_i,\hat{g} \left(x^R_i\right)\right)\right\rbrace, \hspace{0.2in} i =1,2,...,m
 \]
From $(8)$, we learned that observations from the same cluster have been clustered together in the training set. 
And if $g_j = \hat{g} \left(x^R_i\right)$ for observation $i$
\[x^R_i = \mu^R \left(c_i\right) + \varepsilon^R_i = \mu^R_{j} + \varepsilon^R_i\] 
Which implies  
\[x^L_i = \mu^L \left(c_i\right) + \varepsilon^L_i = \mu^L_{j} + \varepsilon^L_i\] 
Hence training set of $\mathcal{C}$ consists of data in the form of 
\[ \left(\mathbf{X}^L_{train},\mathcal{G} \right) = \left\lbrace\left(\mu^L_{j} + \varepsilon^L_i, g_j \right)\right\rbrace\hspace{0.2in} j =1,2,...,K, \hspace{0.1in} i=1,...,m\]
So Let $\hat{\mu}^L_{j} = \frac{\sum x^L_i \cdot I_{c_i=j} }{n^{train}_j}$, $x^L_i  \in \mathbf{X}^L_{train}$ denotes the sample mean of each cluster in $\mathbf{X}^L_{train}$. Define following classifier
	 \[ \mathcal{C}(x^L_i) = \sum^K_{j=1} g_j \cdot I_{h(i) = j} \]
	 where\[h(i) = \underset{j}{\operatorname{argmin}} \hspace{0.05 in}||x^L_i - \hat{\mu}^L_j|| \]
	Note here $x^L_i  \in \mathbf{X}^L_{test}$.\\

Then, above classifier has prediction error $0$. To prove it, we need to show above classifier $\mathcal{C}(x^L_i)$ always gives the same cluster as the one in equation $(8)$, i.e. 
\begin{equation}
\mathcal{C}(x^L_i) = \sum^K_{j=1} g_j \cdot I_{c_i = j}
\end{equation}
Given $\mathcal{D}$, we have
\[||x^L_i-\hat{\mu}^L \left(c_i\right)|| < ||x^L_i-\hat{\mu}^L_{j}||\hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\]
Which means
\[h(i) = \underset{j}{\operatorname{argmin}} \hspace{0.05 in}||x^L_i - \hat{\mu}^L_j|| = c_i \]
$(9)$ follows directly.
\end{proof}
\noindent
If we further assume equal prior are observed in training set, then we can show that the LDA classifier has prediction error $0$, that is
\begin{proposition}
Conditioning on $\mathcal{D}$ and $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$ and $n^{train}_j=\frac{m}{K}$ $\forall j$ in training set, the classifier $\mathcal{C}$ trained using LDA has prediction error $0$ in case of $k = K$.
\end{proposition}
\begin{proof}
Same as in the proof of proposition $4$, given $A_m(K) = \{\hat{\mu}^R_1,\hat{\mu}^R_2,...,\hat{\mu}^R_K \}$, the training set of $\mathcal{C}$ consists of data in the form of 
\[ \left(\mathbf{X}^L_{train},\mathcal{G} \right) = \left\lbrace\left(\mu^L_{j} + \varepsilon^L_i, g_j \right)\right\rbrace\hspace{0.2in} j =1,2,...,K, \hspace{0.1in} i=1,...,m\]
The class-conditional density of $X^L$ in class $\mathcal{G} = g_j$ is $$f_j\left(x^L_i\right) = \mathbb{P}\left( X^L=x^L_i | \mathcal{G} = g_j \right) $$ where $x^L_i=\mu^L(c_i)+\varepsilon^L_i$ and $f_j\left(x^L_i\right)$ has multivariate Gaussian distribution
\[ f_j\left(x^L_i\right) = \frac{1}{(2\pi)^{L/2}|\Sigma^L_j|^{1/2}}e^{-\frac{1}{2}(x^L_i-\mu^L_j)^T(\Sigma^L_j)^{-1}(x^L_i-\mu^L_j)	}\]
LDA assumes a commom covariance matrix $\Sigma^L_j = \Sigma^L \hspace{0.1in} \forall j$.
The posterior probability is 
\[	\mathbb{P}\left( \mathcal{G} = g_j | X^L=x^L_i  \right)  = \frac{f_j\left(x^L_i\right) \pi_j}{\sum^K_{l=1}f_l\left(x^L_i\right) \pi_l}	\]
where $ \pi_j$ is the prior for class $g_j$. In practice, $\pi_j, \mu^L_j$ and $\Sigma^L$ are estimated from data by
 \begin{itemize}
 	\item  $\hat{\pi}_j = \frac{n^{train}_j}{m}$, where $n^{train}_j$ is the number of class-$g_j$ observation in training set and $m$ is the total observations in training set
 	\item  $\hat{\mu}^L_{j} = \frac{\sum x^L_i \cdot I_{c_i=j} }{n^{train}_j}$, $x^L_i  \in \mathbf{X}^L_{train}$
 	\item  $\hat{\Sigma}^L = \frac{1}{m-K} \sum^K_{j=1}\sum_{c_i=j} (x^L_i - \hat{\mu}^L_j )(x^L_i - \hat{\mu}^L_j )^T$
 \end{itemize}
Because $\forall j$ $n^{train}_j=\frac{m}{K}$ , $\hat{\pi}_j = \frac{1}{K}$ $\forall j$. LDA predicts class for each $x^L_i  \in \mathbf{X}^L_{test}$ by maximizing the posterior probability
\begin{align}
     \hat{\mathcal{G}}(x_i) & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} \mathbb{P}\left( \mathcal{G} = g_j | X^L=x^L_i  \right)  \nonumber   \\ 
     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x^L_i\right) \hat{\pi}_j \nonumber \nonumber  \\
   \hat{\pi}_j = \frac{1}{K} \hspace{0.1in}\forall j  \hspace{0.2in} & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in} f_j\left(x^L_i\right) \nonumber  \nonumber  \\    
     & = \underset{g_j}{\operatorname{argmax}}\hspace{0.05 in}  \frac{1}{(2\pi)^{L/2}|\hat{\Sigma}^L|^{1/2}}e^{-\frac{1}{2}(x^L_i-\hat{\mu}^L_j)^T(\hat{\Sigma}^{L})^{-1}(x^L_i-\hat{\mu}^L_j)	} \nonumber  \\  
     &=  \underset{g_j}{\operatorname{argmin}}\hspace{0.05 in} (x^L_i-\hat{\mu}^L_j)^T(\hat{\Sigma}^{L})^{-1}(x^L_i-\hat{\mu}^L_j) \nonumber   \\   \nonumber 
     &= \sum^K_{j=1} g_j \cdot I_{c_i=j}
\end{align}
The last equation is because given $\mathcal{D}$, we have
\[||x^L_i-\hat{\mu}^L \left(c_i\right)|| < ||x^L_i-\hat{\mu}^L_{j}||\hspace{0.2 in}\forall \hspace{0.01 in}j\neq c_i\]
Since it coincides with equation $9$, the proof is complete. 
\end{proof}
\vspace{0.3in}
\noindent
By definition,
\begin{equation}
Error_k = \sum^k_{j=1} \sum_{g_j = \hat{g} \left(x^L_i\right)} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 = \sum^{n-m}_{i=1} ||\hat{\mu}^R \left(g_j\right) - x^R_i||^2, \hspace{0.1in}with\hspace{0.1in} g_j = \hat{g}(x^L_i)
\end{equation}
\noindent
For any parameter $k\leq k_{max}$, $x_i \in \mathbf{X}_{test}$ and $||\hat{\mu}^R \left(g_j\right) - x^R_i||^2$ in $(10)$ are i.i.d with finite expectation
\[
E||\hat{\mu}^R \left(g_j\right) - x^R_i||^2 < \infty
\]
Since $||\hat{\mu}^R \left(g_j\right) - x^R_i||$ is bounded 
\begin{align} \nonumber 

>>>>

\documentclass[11pt]{article}
\usepackage{epsf}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath}
\usepackage{footnote}
\usepackage[bottom]{footmisc}
\usepackage[singlelinecheck=false]{caption}
\usepackage{caption}
\usepackage{amsthm}
\usepackage{enumitem}
\usepackage{graphics}
\usepackage{float}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{lscape}
\usepackage{array}
\textheight=8.7in
\usepackage{setspace}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{longtable}

@@

% operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\CV}{\operatorname{CV}}
\newcommand{\PE}{\operatorname{PE}}
\newcommand{\E}{\operatorname{E}}
\DeclareMathOperator*{\cov}{cov}
\DeclareMathOperator*{\tr}{tr}
\DeclareMathOperator*{\var}{var}

@@

\section{Asymptotic Analysis of $k$-means}

@@

\subsection{Single Cluster}

@@

\subsection{Well-Separated Clusters with Noise}

In this section, simulation is performed to evaluate the performance of our
proposed methods in locating the ``correct" number of clusters. We compare
with a basket of existing methods including Gap statistics
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, CH-index \citep{calinski1974dendrite}, Hartigan
statistics \citep{hartigan1975clustering}, Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Bootstrap stability \citep{fang2012selection} in following simulation
settings. We select $L=R$ and $5-$fold cross-validation in row
($m=\frac{4}{5}n$) as default parameter setting for our proposed Gabriel
method. Note that set $L=R$ corresponding to $2-$fold cross-validation in
column.
\begin{enumerate}

  \item Single cluster in $10$ dimensions --- $200$ observations, each
    observation uniformly distributed over $[0,1]$ in each dimension.

  \item Two clusters in $4$ dimensions --- $50$ i.i.d standard normal
    observations are generated, i.e. generated from multivariate normal
    $\mathcal{N}\left(\boldsymbol\mu,\mathbf{I}\right)$ for each cluster, with
    cluster center $\boldsymbol\mu = (1,0,0,1)$ and $(1,3.5,3.5,1)$
    respectively

  \item Four clusters in $100$ dimensions ---  Each cluster has $100$ or $150$
    i.i.d standard normal observations, with cluster centers randomly
    generated from multivariate normal distribution
    $\mathcal{N}\left(\mathbf{0},0.65^2\mathbf{I}\right)$

  \item Ten clusters in $100$ dimensions ---  Each cluster has $50$ or $100$
    i.i.d standard normal observations, with cluster centers randomly
    generated from multivariate normal distribution
    $\mathcal{N}\left(\mathbf{0},0.72^2\mathbf{I}\right)$

  \item Four log-normal clusters in $16$ dimensions --- For each cluster, $30$
    or $60$ i.i.d centered log-normal observations are generated from
    $ln\mathcal{N}\left(0,0.5^2 \right)$. The cluster centers are randomly
    generated from multivariate normal distribution
    $\mathcal{N}\left(\mathbf{0},1.2^2\mathbf{I}\right)$

  \item[6.] Three exponential clusters in $20$ dimensions with different
    variance --- $40$ observations in each cluster are generated from centered
    exponential distribution ($exp(\lambda)-1/\lambda$ in each dimension),
    with $\boldsymbol\mu$ randomly generated from
    $\mathcal{N}\left(\mathbf{0},19\mathbf{I}\right)$ and $\lambda = 1, 1/2,
    1/5$ respectively.
\end{enumerate}
Note that in setting 2--6, all clusters are well-separated, i.e. no
overlapping. In fact, any simulated clusters with minimum distance less than
1 unit was discarded, so there is clear definition of true number of
clusters. The parameters in setting 2--6 are chosen such that about half of
the random realization were discarded. The idea is borrowed from
\cite{tibshirani2001estimating}.

Besides the number of clusters selected by each algorithm is returned, we also
give the prediction error (PE) for each algorithm in each setting. The PE is
defined by 
\[
  \PE(k)
    =
      \frac{\|\hat{\mathbf{X}}_k- \boldsymbol{\mu} \|_F}
           {PE_{\text{oracle}}}
\]
where
\[
  \PE_{\text{oracle}} = \min_k \|\hat{\mathbf{X}}_k- \boldsymbol{\mu} \|_F
\]
Matrix $\boldsymbol{\mu}$ consists of true centers of observations in data
matrix $\mathbf{X}$, while $\hat{\mathbf{X}}_k$ is obtained by replacing each
observation in $\mathbf{X}$ by its closest cluster center result from applying
K-means on $\mathbf{X}$ with parameter $k$. Here $k$ is the optimal number of
clusters selected by each algorithm on data $\mathbf{X}$. This is only
possible on simulated data because otherwise one would never know the true
center matrix~$\boldsymbol{\mu}$.

The oracle selected number for each data $\mathbf{X}$ is also given on the
table, which is defined by 
\[
  K_{\text{oracle}}
    =
      \argmin_{k} \|\hat{\mathbf{X}}_k - \boldsymbol{\mu} \|_F
\]

@@

\begin{center}
{%\footnotesize
\begin{longtable}{lcccccccccccccccc}
\caption{\label{table1} Simulation Results}\\
\toprule
\phantom{Pred.~Strength}
& \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15
  & \textsc{na} \\
\midrule
\endfirsthead
\multicolumn{17}{c}%
{\tablename\ \thetable\ -- \textit{Continued from previous page}} \\
\toprule
\phantom{Pred.~Strength}
  & \multicolumn{16}{c}{Estimated Number of Clusters} \\
\cmidrule(lr){2-17}
Method
  & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15
  & \textsc{na} \\
\midrule
\endhead
%\bottomrule
\multicolumn{17}{r}{\textit{Continued on next page}} \\
\endfoot
\bottomrule
\multicolumn{17}{l}{
{\footnotesize {In Setting~1, ``\textendash'' means the method can not be used
with parameter $k=1$}.}}
\endlastfoot
\input{demo/bench/setting1/results.tex}
\midrule
\input{demo/bench/setting2/results.tex}
\midrule
\input{demo/bench/setting3/results.tex}
\midrule
\input{demo/bench/setting4/results.tex}
\midrule
\input{demo/bench/setting5/results.tex}
\midrule
\input{demo/bench/setting6/results.tex}
\end{longtable}
}
\end{center}

\noindent
All methods are used with their default parameter setting except for Gap
statistics. We selected the $k$  corresponds to the global maximum of gap
statistics, rather than the default method which selects the smallest $k$ such
that the gap statistics is not more than 1 standard error away from the first
local maximum. This is because among all the possible options, the global
maximum criteria gives the best result in our simulation settings.

The ``\textendash'' sign in Setting~1 is because for CH-index, Hartigen and
Bootstrap stability, the minimum $k$ they can pick is $2$. So they are not
included for comparison since the true $k=1$. Setting $2-4$ have Gaussian
distribution, where the Gaussian mixture model-based clustering suppose to
perform the best. It is only the case for setting $2$. In setting $3$ and $4$,
it has a very poor performance even if the underlying distribution is indeed
Gaussian mixture model. It highlights the difficulties that high dimension
data could cause. Other methods (Gap statistics, CH-index, Hartigan, Jump
method, Prediction strength and Boostrap stability) also have problems in
these two high-dimension settings as we could not find any one with reasonable
performance in terms of finding the correct $k$ in both cases. In terms of PE,
Boostrap stability and Gap statistics have reasonable performance, but come
with the price of picking larger $k$. Our proposed Gabriel method (also Wold
CV method) clearly stands out from the basket of existing methods using either
measure under the high-dimension setting $3$ and $4$.

Setting $5$ and $6$ corresponding to the situation where data are heavy-tailed
instead of normal distributed, coupled with unequal number of observations or
unequal variance for different clusters. From Table \ref{table1}, we can see
that our proposed Gabriel method is robust and has superior performance
compare with other methods. Because one can hardly tell the underlying
distribution of data in practice, the resilience to non-Gaussian data gives
the Gabriel method a clear edge.

\section{Real data application}

We also applied our proposed method to three real world data sets obtained
from the University of California Irvine machine learning repository. The
first and third data sets are selected because there are clear number of
clusters in those two data sets. The second data set is used as a benchmark
data set since it was widely used in literature.

The first one is congress voting data which consists of voting records of
$98$th United States Congress, $2$nd session. This data set includes votes for
each of the U.S. House of Representatives Congressmen on the $16$ key votes
identified by the \textit{CQA} (Congressional Quarterly Almanac). For each
vote, each Congressman either vote positively ``yea" (voted for/paired
for/announced for),  negatively ``nay'' (voted against/paired
against/announced against) or position unknown ``?". We took out those records
contain ``?" before comparing each algorithm. It results in $232$ remaining
records, with $124$ democrat and $108$ republican.
